---
title: "new analysis with annotated mz only"
author: "Nick Bartelo"
date: "2024-08-22"
output: html_document
---

First we install some packages.

```{r, warning=FALSE, message = FALSE}
library(magrittr)
library(dplyr)
library(tidyverse)
library(data.table)
library(car)
library(stats)
library(ggplot2)
library(fable)
library(ggfortify)
#devtools::install_github(repo="krumsieklab/maplet", subdir="maplet")
library(maplet)
```

Below we import the data, create metadata for the samples, drop the column used to make the metadata, transpose the data so features are rows and samples are columns, replace 0 values with missing values, and make it a matrix.


```{r}
# Import data
# Define path to data
path_to_folder_with_data <- "/Users/nicholasbartelo/Documents/Zhao-Lab-AMR-Paper-main/Data Preprocessing/Input File/"

# Read in data
data <- read.csv(
  file.path(path_to_folder_with_data, "Raw data export of all peaks_no normalization080522_for_maplet.csv"),
  row.names = 2, header = TRUE
)

# Make into CRE and CSE only
data$Group <- ifelse(data$Group %like% c("CRE"), "CRE", "CSE")

# Import species info
species_info <- read.csv(
  file.path(path_to_folder_with_data, "Raw_data_with_organism_info.csv"))
# Convert rownames of df1 into a column
data$Sample <- rownames(data)
# Merge the dataframes by the 'Sample' column
data_with_organism <- merge(data, species_info[,c("Sample", "Organism")], by = "Sample", all = TRUE)
# Make rownames the sample names
rownames(data_with_organism) <- data_with_organism$Sample

data_with_organism$date_measured <- ifelse(grepl("421", data_with_organism$Sample), "apr21", ifelse(grepl("425", data_with_organism$Sample), "apr25", "apr27"))

# Make list of species to keep
species_to_keep <- c('Klebsiella pneumoniae', 'Enterobacter cloacae', 'Serratia marcescens', 'Proteus mirabilis', 'Klebsiella oxytoca', 'Citrobacter freundii', 'Klebsiella aerogenes', 'Escherichia coli')
# Keep only the desired species
data_with_organism_filtered <- data_with_organism[data_with_organism$Organism %in% species_to_keep, ]

# Create dataframe for metadata for SE object
column_identifiers <- data_with_organism_filtered[, c("Group", "Organism", "date_measured"), drop=FALSE]
# Remove unwanted columns
data_with_organism_filtered$Sample <- NULL
data_with_organism_filtered$Group <- NULL
data_with_organism_filtered$Organism <- NULL
data_with_organism_filtered$date_measured <- NULL

# Transpose dataframe
data_with_organism_filtered = t(data_with_organism_filtered)
# Set all 0 values to missing
data_with_organism_filtered[data_with_organism_filtered == 0] <- NA
# Make dataframe into matrix
data_with_organism_filtered <- as.matrix(data_with_organism_filtered)
```

Below we import the annotated feature matrix and keep only the overlapping metabolites.

```{r}
# Import data
annotations_mz <- read.csv(
  file.path(path_to_folder_with_data, "mapped_data_under_01_tolerance.csv"), header = TRUE)
# Remove X from rownames
rownames(data_with_organism_filtered) <- sub("^X", "", rownames(data_with_organism_filtered))
# Keep only rownames that are in annotated features
data_with_organism_filtered_mz_subset <- data_with_organism_filtered[rownames(data_with_organism_filtered) %in% annotations_mz$mz, ]

# Reorder annotation names to match metabolite abundance data
#annotations_mz_reordered <- annotations_mz[match(rownames(data_with_organism_filtered_mz_subset), annotations_mz$mz), ]
# Merge compound names to data
#rownames(data_with_organism_filtered_mz_subset) <- annotations_mz_reordered$Compound.name
```


Below we make our data into a summarized experiment object.

```{r}
raw_SE <- SummarizedExperiment(assays=list(counts=data_with_organism_filtered_mz_subset), colData = column_identifiers)
```

Below we use maplet to process the data.

```{r}
processed_SE <- raw_SE %>%
  mt_pre_filter_missingness(feat_max = 0.2) %>% # Filter out features with more than 20% missing
  mt_pre_filter_missingness(samp_max = 0.1) %>%  # Filter out samples with more than 10% missing
  mt_pre_norm_quot() %>% # Quotient Normalization
  mt_pre_trans_log() %>% # Log transform
  mt_pre_trans_scale() %>% # Scale data (z-scores)
  mt_pre_outlier_to_na() %>% # Detect outliers (any values that lie outside the 0.025 and 0.975 regions when assuming a normal distribution)  and set to NA
  mt_pre_impute_knn() # Impute missing values with knn imputation

processed_SE <- processed_SE %>% 
  mt_plots_pca(color = Group)

processed_SE <- processed_SE %>% 
  mt_plots_pca(color = Organism)

processed_SE <- processed_SE %>% 
  mt_plots_pca(color = date_measured)
```

Below we look at the PCA plots.

```{r}
processed_SE@metadata$results$`plots_pca.e9610723-f734-4ddc-a326-f2ce24f16d93`

processed_SE@metadata$results$`plots_pca.1792d0a3-7047-4460-8657-899b67004c4e`

processed_SE@metadata$results$`plots_pca.1e9f86ef-39e1-4d2f-8840-53748e7224ca`
```

Below we create an assay of the data and create a PCA plot to combine the phenotype and organism data.

```{r}
# Make dataframe of processed data
df_processed <- as.data.frame(t(assay(processed_SE)))

# Compute PCA
pca_result <- prcomp(df_processed)

# Calculate the percentage of explained variance for each PC
explained_variance <- (pca_result$sdev^2 / sum(pca_result$sdev^2)) * 100

# Extract the explained variance for PC1 and PC2
pc1_var <- round(explained_variance[1], 2)  # Variance explained by PC1
pc2_var <- round(explained_variance[2], 2)  # Variance explained by PC2

# Prepare metadata for plotting
processed_se_group <- processed_SE@colData$Group
processed_se_organism <- processed_SE@colData$Organism
processed_se_metadata <- data.frame(processed_se_group, processed_se_organism)

# Create a data frame with PCA results and metadata
pca_data <- as.data.frame(pca_result$x)
pca_data <- cbind(pca_data, processed_se_metadata)

# Plot PCA results using ggplot2
final_plot_before_batch_correction_organism_phenotype <- ggplot(pca_data, aes(x = PC1, y = PC2, color = processed_se_organism, shape = processed_se_group)) +
  geom_point(size = 3) +  # Adjust point size if necessary
  theme_minimal() +       # Minimal theme (white background)
  theme(panel.border = element_blank(),
        panel.background = element_rect(fill = "white"), # Ensure background is white
        plot.background = element_rect(fill = "white", colour = NA),
  
   # Increase label sizes
        axis.title.x = element_text(size = 16), # X-axis label size
        axis.title.y = element_text(size = 16), # Y-axis label size
        legend.title = element_text(size = 16), # Legend title size
        legend.text = element_text(size = 15),  # Legend text size
        axis.text = element_text(size = 15)     # Axis tick mark text size
        ) +
  
  # Manually set color labels with one label italicized
  scale_color_manual(values = c("blue", "brown", "gray", "black", "orange", "pink", "yellow", "violet"),
                     labels = c(expression(italic("Citrobacter freundii")), expression(italic("Enterobacter cloacae")), expression(italic("Escherichia coli")), expression(italic("Klebsiella aerogenes")), expression(italic("Klebsiella oxytoca")), expression(italic("Klebsiella pneumoniae")), expression(italic("Proteus mirabilis")), expression(italic("Serratia marcescens")))) +
  
  # Change the legend titles and add the explained variance to the axis labels
  labs(
    x = paste0("Principal Component 1 (", pc1_var, "% variance)"),
    y = paste0("Principal Component 2 (", pc2_var, "% variance)"),
    color = "Organism",
    shape = "Phenotype"
  ) +
  
  # Ensure equal scaling on both axes
  coord_fixed()
```


Below we create a PCA plot to show the batch data.

```{r}
# Make dataframe of processed data
df_processed <- as.data.frame(t(assay(processed_SE)))
# Compute PCA
pca_result <- prcomp(df_processed)
# Calculate the percentage of explained variance for each PC
explained_variance <- (pca_result$sdev^2 / sum(pca_result$sdev^2)) * 100
# Extract the explained variance for PC1 and PC2
pc1_var <- round(explained_variance[1], 2)  # Variance explained by PC1
pc2_var <- round(explained_variance[2], 2)  # Variance explained by PC2

processed_se_date <- processed_SE@colData$date_measured
processed_se_date_df <- data.frame(processed_se_date)

# Create a data frame with the PCA results and original data
pca_data <- as.data.frame(pca_result$x)
pca_data <- cbind(pca_data, processed_se_date_df)

# Plot PCA results using ggplot2
final_plot_before_batch_correction_date_measured <- ggplot(pca_data, aes(x = PC1, y = PC2, color = processed_se_date)) +
  geom_point(size = 3) +  # Adjust point size if necessary
  theme_minimal() +       # Minimal theme (white background)
  theme(panel.border = element_blank(),
        panel.background = element_rect(fill = "white"), # Ensure background is white
        plot.background = element_rect(fill = "white", colour = NA),
  
   # Increase label sizes
        axis.title.x = element_text(size = 16), # X-axis label size
        axis.title.y = element_text(size = 16), # Y-axis label size
        legend.title = element_text(size = 20), # Legend title size
        legend.text = element_text(size = 20),  # Legend text size
        axis.text = element_text(size = 15)     # Axis tick mark text size
        ) +
  
    # Manually set color labels with one label italicized
  scale_color_manual(values = c("purple", "cyan", "lightgreen"), 
                     labels = c("April 21", "April 25", "April 27")) +
  
  # Change the legend titles and add the explained variance to the axis labels
  labs(
    x = paste0("Principal Component 1 (", pc1_var, "% variance)"),
    y = paste0("Principal Component 2 (", pc2_var, "% variance)"),
    color = "Measurement Date"
  ) +
  
  # Ensure equal scaling on both axes
  coord_fixed()
```

Below we create a PCA plot to show the batch data but we combine the April 25 and April 27 variable into one.

```{r}
# Make dataframe of processed data
df_processed <- as.data.frame(t(assay(processed_SE)))
# Compute PCA
pca_result <- prcomp(df_processed)
# Calculate the percentage of explained variance for each PC
explained_variance <- (pca_result$sdev^2 / sum(pca_result$sdev^2)) * 100
# Extract the explained variance for PC1 and PC2
pc1_var <- round(explained_variance[1], 2)  # Variance explained by PC1
pc2_var <- round(explained_variance[2], 2)  # Variance explained by PC2

processed_se_date <- processed_SE@colData$date_measured
# Combine two dates
processed_se_date_two_batch <- ifelse(processed_se_date == "apr21", "April 21", "April 25/27")
processed_se_date_two_batch_df <- data.frame(processed_se_date_two_batch)

# Create a data frame with the PCA results and original data
pca_data <- as.data.frame(pca_result$x)
pca_data <- cbind(pca_data, processed_se_date_two_batch_df)

# Plot PCA results using ggplot2
final_plot_before_batch_correction_date_measured <- ggplot(pca_data, aes(x = PC1, y = PC2, color = processed_se_date_two_batch)) +
  geom_point(size = 3) +  # Adjust point size if necessary
  theme_minimal() +       # Minimal theme (white background)
  theme(panel.border = element_blank(),
        panel.background = element_rect(fill = "white"), # Ensure background is white
        plot.background = element_rect(fill = "white", colour = NA),
  
   # Increase label sizes
        axis.title.x = element_text(size = 16), # X-axis label size
        axis.title.y = element_text(size = 16), # Y-axis label size
        legend.title = element_text(size = 20), # Legend title size
        legend.text = element_text(size = 20),  # Legend text size
        axis.text = element_text(size = 15)     # Axis tick mark text size
        ) +
  
    # Manually set color labels with one label italicized
  scale_color_manual(values = c("purple", "lightgreen"), 
                     labels = c("April 21", "April 25/27")) +
  
  # Change the legend titles and add the explained variance to the axis labels
  labs(
    x = paste0("Principal Component 1 (", pc1_var, "% variance)"),
    y = paste0("Principal Component 2 (", pc2_var, "% variance)"),
    color = "Measurement Date"
  ) +
  
  # Ensure equal scaling on both axes
  coord_fixed()
```

### Statistical testing

Below we run an ANOVA to test if the distributions of metabolites are different before batch correction. This gives 305/512 ~ 60% significant metabolites being different before batch correction between batches. 

```{r}
# Make new dataframe
df_processed_no_correction <- as.data.frame(t(assay(processed_SE)))
# Add columns from metadata
df_processed_no_correction$Group <- processed_SE@colData$Group
df_processed_no_correction$Organism <- processed_SE@colData$Organism
df_processed_no_correction$date_measured <- processed_SE@colData$date_measured
# Add new batches
df_processed_no_correction$batch_measured <- ifelse(df_processed_no_correction$date_measured %like% c("apr21"), "apr21", "apr25_27")

# Make list for pvalues to be added to
aov_pval_list_before_batch_correction <- list()

final_statistics <- lapply(1:(ncol(df_processed_no_correction) - 4), function(i){
    anova_df <- data.frame(metabolite=df_processed_no_correction[,i], df_processed_no_correction$batch_measured)
    aov_test <- aov(metabolite ~ df_processed_no_correction$batch_measured, data=anova_df)
    aov_pval <- summary(aov_test)[[1]][["Pr(>F)"]][1]
    aov_pval_list_before_batch_correction[[length(aov_pval_list_before_batch_correction) + 1]] <- aov_pval
}) 
  
final_aov_dataframe <- data.frame(colnames(df_processed_no_correction)[1:(ncol(df_processed_no_correction) - 4)], unlist(final_statistics))

significant_aov_dataframe <- final_aov_dataframe[final_aov_dataframe$unlist.final_statistics. < 0.05, ]

# Turn all pvalues into adjusted pvalues by applying FDR correction which multiples all values by the number of metabolites
final_aov_dataframe_with_correction <- final_aov_dataframe %>% dplyr::mutate(group_padj = p.adjust(final_aov_dataframe$unlist.final_statistics., method = 'fdr'))

significant_aov_dataframe_with_correction <- final_aov_dataframe_with_correction[final_aov_dataframe_with_correction$group_padj < 0.05, ]
```

# Batch Correction Using Combat

```{r}
# Read in data
data <- read.csv(
  file.path(path_to_folder_with_data, "Raw data export of all peaks_no normalization080522_for_maplet.csv"),
  row.names = 2, header = TRUE
)
# Make into CRE and CSE only
data$Group <- ifelse(data$Group %like% c("CRE"), "CRE", "CSE")

# Import species info
species_info <- read.csv(
  file.path(path_to_folder_with_data, "Raw_data_with_organism_info.csv"))
# Convert rownames of df1 into a column
data$Sample <- rownames(data)
# Merge the dataframes by the 'Sample' column
data_with_organism <- merge(data, species_info[,c("Sample", "Organism")], by = "Sample", all = TRUE)
# Make rownames the sample names
rownames(data_with_organism) <- data_with_organism$Sample

data_with_organism$date_measured <- ifelse(grepl("421", data_with_organism$Sample), "apr21", ifelse(grepl("425", data_with_organism$Sample), "apr25", "apr27"))
# Add new batches
data_with_organism$batch_measured <- ifelse(data_with_organism$date_measured %like% c("apr21"), "apr21", "apr25_27")

# Make list of species to keep
species_to_keep <- c('Klebsiella pneumoniae', 'Enterobacter cloacae', 'Serratia marcescens', 'Proteus mirabilis', 'Klebsiella oxytoca', 'Citrobacter freundii', 'Klebsiella aerogenes', 'Escherichia coli')
# Keep only the desired species
data_with_organism_filtered <- data_with_organism[data_with_organism$Organism %in% species_to_keep, ]

# Create dataframe for metadata for SE object
column_identifiers <- data_with_organism_filtered[, c("Group", "Organism", "date_measured", "batch_measured"), drop=FALSE]
# Remove unwanted columns
data_with_organism_filtered$Sample <- NULL
data_with_organism_filtered$Group <- NULL
data_with_organism_filtered$Organism <- NULL
data_with_organism_filtered$date_measured <- NULL
data_with_organism_filtered$batch_measured <- NULL


# Transpose dataframe
data_with_organism_filtered = t(data_with_organism_filtered)
# Set all 0 values to missing
data_with_organism_filtered[data_with_organism_filtered == 0] <- NA
# Make dataframe into matrix
data_with_organism_filtered <- as.matrix(data_with_organism_filtered)
```

Below we import the annotated feature matrix and keep only the overlapping metabolites.

```{r}
# Import data
annotations_mz <- read.csv(
  file.path(path_to_folder_with_data, "mapped_data_under_01_tolerance.csv"), header = TRUE)
# Remove X from rownames
rownames(data_with_organism_filtered) <- sub("^X", "", rownames(data_with_organism_filtered))
# Keep only rownames that are in annotated features
data_with_organism_filtered_mz_subset <- data_with_organism_filtered[rownames(data_with_organism_filtered) %in% annotations_mz$mz, ]

# Reorder annotation names to match metabolite abundance data
#annotations_mz_reordered <- annotations_mz[match(rownames(data_with_organism_filtered_mz_subset), annotations_mz$mz), ]
# Merge compound names to data
#rownames(data_with_organism_filtered_mz_subset) <- annotations_mz_reordered$Compound.name
```


Below we make our data into a summarized experiment object.

```{r}
raw_SE <- SummarizedExperiment(assays=list(counts=data_with_organism_filtered_mz_subset), colData = column_identifiers)
```

Below we use maplet to process the data.

```{r}
processed_SE <- raw_SE %>%
  mt_pre_filter_missingness(feat_max = 0.2) %>% # Filter out features with more than 20% missing
  mt_pre_filter_missingness(samp_max = 0.1) %>%  # Filter out samples with more than 10% missing
  mt_pre_norm_quot() %>% # Quotient Normalization
  mt_pre_trans_log() %>% # Log transform
  mt_pre_trans_scale() %>% # Scale data (z-scores)
  mt_pre_outlier_to_na() %>% # Detect outliers (any values that lie outside the 0.025 and 0.975 regions when assuming a normal distribution)  and set to NA
  mt_pre_impute_knn() %>% # Impute missing values with knn imputation
  mt_pre_batch_combat(batch_col = "batch_measured") # combat batch correction NEEDS TO BE DONE AFTER MISSING VALUES TAKEN CARE OF

processed_SE <- processed_SE %>% 
  mt_plots_pca(color = processed_SE@colData$Group)

processed_SE <- processed_SE %>% 
  mt_plots_pca(color = processed_SE@colData$Organism)

processed_SE <- processed_SE %>% 
  mt_plots_pca(color = processed_SE@colData$date_measured)

processed_SE <- processed_SE %>% 
  mt_plots_pca(color = processed_SE@colData$batch_measured)
```

Below we look at the PCA plots.

```{r}
processed_SE@metadata$results$`plots_pca.23eb6ebf-8a11-441c-91f5-f7798ae5f08b`

processed_SE@metadata$results$`plots_pca.1f0039fe-d569-412d-ae4f-c78ca2eef6f0`

processed_SE@metadata$results$`plots_pca.8e8f64d2-6224-4aa1-82ee-137f5774d8b0`

processed_SE@metadata$results$`plots_pca.2b2595cf-9e1b-441f-9901-da4cec91e338`
```

Below we create an assay of the data and create a PCA plot to combine the phenotype and organism data.

```{r}
# Make dataframe of processed data
df_processed <- as.data.frame(t(assay(processed_SE)))

# Compute PCA
pca_result <- prcomp(df_processed)

# Calculate the percentage of explained variance for each PC
explained_variance <- (pca_result$sdev^2 / sum(pca_result$sdev^2)) * 100

# Extract the explained variance for PC1 and PC2
pc1_var <- round(explained_variance[1], 2)  # Variance explained by PC1
pc2_var <- round(explained_variance[2], 2)  # Variance explained by PC2

# Prepare metadata for plotting
processed_se_group <- processed_SE@colData$Group
processed_se_organism <- processed_SE@colData$Organism
processed_se_metadata <- data.frame(processed_se_group, processed_se_organism)

# Create a data frame with PCA results and metadata
pca_data <- as.data.frame(pca_result$x)
pca_data <- cbind(pca_data, processed_se_metadata)

# Plot PCA results using ggplot2
final_plot_after_batch_correction_organism_phenotype <- ggplot(pca_data, aes(x = PC1, y = PC2, color = processed_se_organism, shape = processed_se_group)) +
  geom_point(size = 3) +  # Adjust point size if necessary
  theme_minimal() +       # Minimal theme (white background)
  theme(panel.border = element_blank(),
        panel.background = element_rect(fill = "white"), # Ensure background is white
        plot.background = element_rect(fill = "white", colour = NA),
  
   # Increase label sizes
        axis.title.x = element_text(size = 15), # X-axis label size
        axis.title.y = element_text(size = 15), # Y-axis label size
        legend.title = element_text(size = 16), # Legend title size
        legend.text = element_text(size = 15),  # Legend text size
        axis.text = element_text(size = 15)     # Axis tick mark text size
        ) +
  
  # Manually set color labels with one label italicized
  scale_color_manual(values = c("blue", "brown", "gray", "black", "orange", "pink", "yellow", "violet"),
                     labels = c(expression(italic("Citrobacter freundii")), expression(italic("Enterobacter cloacae")), expression(italic("Escherichia coli")), expression(italic("Klebsiella aerogenes")), expression(italic("Klebsiella oxytoca")), expression(italic("Klebsiella pneumoniae")), expression(italic("Proteus mirabilis")), expression(italic("Serratia marcescens")))) +
  
  # Change the legend titles and add the explained variance to the axis labels
  labs(
    x = paste0("Principal Component 1 (", pc1_var, "% variance)"),
    y = paste0("Principal Component 2 (", pc2_var, "% variance)"),
    color = "Organism",
    shape = "Phenotype"
  ) +
  # Ensure equal scaling on both axes
  coord_fixed(ratio = 1.5)
```


Below we create a PCA plot to show the batch data.

```{r}
# Make dataframe of processed data
df_processed <- as.data.frame(t(assay(processed_SE)))
# Compute PCA
pca_result <- prcomp(df_processed)
# Calculate the percentage of explained variance for each PC
explained_variance <- (pca_result$sdev^2 / sum(pca_result$sdev^2)) * 100
# Extract the explained variance for PC1 and PC2
pc1_var <- round(explained_variance[1], 2)  # Variance explained by PC1
pc2_var <- round(explained_variance[2], 2)  # Variance explained by PC2

processed_se_date <- processed_SE@colData$date_measured
processed_se_date_df <- data.frame(processed_se_date)

# Create a data frame with the PCA results and original data
pca_data <- as.data.frame(pca_result$x)
pca_data <- cbind(pca_data, processed_se_date_df)

# Plot PCA results using ggplot2
final_plot_after_batch_correction_date_measured <- ggplot(pca_data, aes(x = PC1, y = PC2, color = processed_se_date)) +
  geom_point(size = 3) +  # Adjust point size if necessary
  theme_minimal() +       # Minimal theme (white background)
  theme(panel.border = element_blank(),
        panel.background = element_rect(fill = "white"), # Ensure background is white
        plot.background = element_rect(fill = "white", colour = NA),
  
   # Increase label sizes
        axis.title.x = element_text(size = 15), # X-axis label size
        axis.title.y = element_text(size = 15), # Y-axis label size
        legend.title = element_text(size = 20), # Legend title size
        legend.text = element_text(size = 20),  # Legend text size
        axis.text = element_text(size = 15)     # Axis tick mark text size
        ) +
  
    # Manually set color labels with one label italicized
  scale_color_manual(values = c("purple", "cyan", "lightgreen"), 
                     labels = c("April 21", "April 25", "April 27")) +
  
  # Change the legend titles and add the explained variance to the axis labels
  labs(
    x = paste0("Principal Component 1 (", pc1_var, "% variance)"),
    y = paste0("Principal Component 2 (", pc2_var, "% variance)"),
    color = "Measurement Date"
  ) +
  # Ensure equal scaling on both axes
  coord_fixed(ratio = 1.5)
```

Below we create a PCA plot to show the batch data but we combine the April 25 and April 27 variable into one.

```{r}
# Make dataframe of processed data
df_processed <- as.data.frame(t(assay(processed_SE)))
# Compute PCA
pca_result <- prcomp(df_processed)
# Calculate the percentage of explained variance for each PC
explained_variance <- (pca_result$sdev^2 / sum(pca_result$sdev^2)) * 100
# Extract the explained variance for PC1 and PC2
pc1_var <- round(explained_variance[1], 2)  # Variance explained by PC1
pc2_var <- round(explained_variance[2], 2)  # Variance explained by PC2

processed_se_date <- processed_SE@colData$date_measured
# Combine two dates
processed_se_date_two_batch <- ifelse(processed_se_date == "apr21", "April 21", "April 25/27")
processed_se_date_two_batch_df <- data.frame(processed_se_date_two_batch)

# Create a data frame with the PCA results and original data
pca_data <- as.data.frame(pca_result$x)
pca_data <- cbind(pca_data, processed_se_date_two_batch_df)

# Plot PCA results using ggplot2
final_plot_after_batch_correction_date_measured <- ggplot(pca_data, aes(x = PC1, y = PC2, color = processed_se_date_two_batch)) +
  geom_point(size = 3) +  # Adjust point size if necessary
  theme_minimal() +       # Minimal theme (white background)
  theme(panel.border = element_blank(),
        panel.background = element_rect(fill = "white"), # Ensure background is white
        plot.background = element_rect(fill = "white", colour = NA),
  
   # Increase label sizes
        axis.title.x = element_text(size = 15), # X-axis label size
        axis.title.y = element_text(size = 15), # Y-axis label size
        legend.title = element_text(size = 20), # Legend title size
        legend.text = element_text(size = 20),  # Legend text size
        axis.text = element_text(size = 15)     # Axis tick mark text size
        ) +
  
    # Manually set color labels with one label italicized
  scale_color_manual(values = c("purple", "lightgreen"), 
                     labels = c("April 21", "April 25/27")) +
  
  # Change the legend titles and add the explained variance to the axis labels
  labs(
    x = paste0("Principal Component 1 (", pc1_var, "% variance)"),
    y = paste0("Principal Component 2 (", pc2_var, "% variance)"),
    color = "Measurement Date"
  ) +
  # Ensure equal scaling on both axes
  coord_fixed(ratio = 1.5)
```

### Statistical testing

Below we run an ANOVA to test if the distributions of metabolites are different after batch correction. This gives 0 significant metabolites being different after batch correction between batches so we can safely say there is no difference between batches! 

```{r}
# Make new dataframe
df_processed_with_correction_after_combat <- as.data.frame(t(assay(processed_SE)))
# Add columns from metadata
df_processed_with_correction_after_combat$Group <- processed_SE@colData$Group
df_processed_with_correction_after_combat$Organism <- processed_SE@colData$Organism
df_processed_with_correction_after_combat$date_measured <- processed_SE@colData$date_measured
# Add new batches
df_processed_with_correction_after_combat$batch_measured <- ifelse(df_processed_with_correction_after_combat$date_measured %like% c("apr21"), "apr21", "apr25_27")

# Make list for pvalues to be added to
aov_pval_list_after_combat <- list()

final_statistics_after_combat <- lapply(1:(ncol(df_processed_with_correction_after_combat) - 4), function(i){
    anova_df <- data.frame(metabolite=df_processed_with_correction_after_combat[,i], df_processed_with_correction_after_combat$batch_measured)
    aov_test <- aov(metabolite ~ df_processed_with_correction_after_combat$batch_measured, data=anova_df)
    aov_pval <- summary(aov_test)[[1]][["Pr(>F)"]][1]
    aov_pval_list_after_combat[[length(aov_pval_list_after_combat) + 1]] <- aov_pval
}) 
  
final_aov_dataframe_after_combat <- data.frame(colnames(df_processed_with_correction_after_combat)[1:(ncol(df_processed_with_correction_after_combat) - 4)], unlist(final_statistics_after_combat))

significant_aov_dataframe_after_combat <- final_aov_dataframe_after_combat[final_aov_dataframe_after_combat$unlist.final_statistics_after_combat. < 0.05, ]

# Turn all pvalues into adjusted pvalues by applying FDR correction which multiples all values by the number of metabolites
final_aov_dataframe_with_correction_after_combat <- final_aov_dataframe_after_combat %>% dplyr::mutate(group_padj_after_combat = p.adjust(final_aov_dataframe_after_combat$unlist.final_statistics_after_combat., method = 'fdr'))

significant_aov_dataframe_with_correction_after_combat <- final_aov_dataframe_with_correction_after_combat[final_aov_dataframe_with_correction_after_combat$group_padj_after_combat < 0.05, ]
```

Below we combine the results of the before and after ANOVA results and save them as a supplementary table for the paper.

```{r}
# Merge by the first column of each dataframe
# Rename columns to specify before or after batch correction and give same name to first column
colnames(final_aov_dataframe_with_correction) <- c("Ion", "Pval_no_batch_correction", "FDR_Pval_no_batch_correction")

colnames(final_aov_dataframe_with_correction_after_combat) <- c("Ion", "Pval_with_batch_correction", "FDR_Pval_with_batch_correction")

merged_df_pvals <- merge(final_aov_dataframe_with_correction, final_aov_dataframe_with_correction_after_combat, by = "Ion")
```


Below we create a new column to specify the batches as apr21 vs apr25 and apr27 since this is how the data separates shown by the image above.

```{r}
# Make new dataframe
df_processed <- as.data.frame(t(assay(processed_SE)))
# Add columns from metadata
df_processed$Group <- processed_SE@colData$Group
df_processed$Organism <- processed_SE@colData$Organism
df_processed$date_measured <- processed_SE@colData$date_measured
df_processed$batch_measured <- processed_SE@colData$batch_measured
# Find the number of samples of each species that are remaining.
counts_organism <- table(df_processed$Organism)
# Display the result
print(counts_organism)
```

# Redoing everything without standardizing the data

```{r}
# Read in data
data <- read.csv(
  file.path(path_to_folder_with_data, "Raw data export of all peaks_no normalization080522_for_maplet.csv"),
  row.names = 2, header = TRUE
)
# Make into CRE and CSE only
data$Group <- ifelse(data$Group %like% c("CRE"), "CRE", "CSE")

# Import species info
species_info <- read.csv(
  file.path(path_to_folder_with_data, "Raw_data_with_organism_info.csv"))
# Convert rownames of df1 into a column
data$Sample <- rownames(data)
# Merge the dataframes by the 'Sample' column
data_with_organism <- merge(data, species_info[,c("Sample", "Organism")], by = "Sample", all = TRUE)
# Make rownames the sample names
rownames(data_with_organism) <- data_with_organism$Sample

data_with_organism$date_measured <- ifelse(grepl("421", data_with_organism$Sample), "apr21", ifelse(grepl("425", data_with_organism$Sample), "apr25", "apr27"))
# Add new batches
data_with_organism$batch_measured <- ifelse(data_with_organism$date_measured %like% c("apr21"), "apr21", "apr25_27")

# Make list of species to keep
species_to_keep <- c('Klebsiella pneumoniae', 'Enterobacter cloacae', 'Serratia marcescens', 'Proteus mirabilis', 'Klebsiella oxytoca', 'Citrobacter freundii', 'Klebsiella aerogenes', 'Escherichia coli')
# Keep only the desired species
data_with_organism_filtered <- data_with_organism[data_with_organism$Organism %in% species_to_keep, ]

# Create dataframe for metadata for SE object
column_identifiers <- data_with_organism_filtered[, c("Group", "Organism", "date_measured", "batch_measured"), drop=FALSE]
# Remove unwanted columns
data_with_organism_filtered$Sample <- NULL
data_with_organism_filtered$Group <- NULL
data_with_organism_filtered$Organism <- NULL
data_with_organism_filtered$date_measured <- NULL
data_with_organism_filtered$batch_measured <- NULL


# Transpose dataframe
data_with_organism_filtered = t(data_with_organism_filtered)
# Set all 0 values to missing
data_with_organism_filtered[data_with_organism_filtered == 0] <- NA
# Make dataframe into matrix
data_with_organism_filtered <- as.matrix(data_with_organism_filtered)
```

Below we import the annotated feature matrix and keep only the overlapping metabolites.

```{r}
# Import data
annotations_mz <- read.csv(
  file.path(path_to_folder_with_data, "mapped_data_under_01_tolerance.csv"), header = TRUE)
# Remove X from rownames
rownames(data_with_organism_filtered) <- sub("^X", "", rownames(data_with_organism_filtered))
# Keep only rownames that are in annotated features
data_with_organism_filtered_mz_subset <- data_with_organism_filtered[rownames(data_with_organism_filtered) %in% annotations_mz$mz, ]

# Reorder annotation names to match metabolite abundance data
#annotations_mz_reordered <- annotations_mz[match(rownames(data_with_organism_filtered_mz_subset), annotations_mz$mz), ]
# Merge compound names to data
#rownames(data_with_organism_filtered_mz_subset) <- annotations_mz_reordered$Compound.name
```


Below we make our data into a summarized experiment object.

```{r}
raw_SE <- SummarizedExperiment(assays=list(counts=data_with_organism_filtered_mz_subset), colData = column_identifiers)
```

Below we use maplet to process the data.

```{r}
processed_SE_not_standardized <- raw_SE %>%
  mt_pre_filter_missingness(feat_max = 0.2) %>% # Filter out features with more than 20% missing
  mt_pre_filter_missingness(samp_max = 0.1) %>%  # Filter out samples with more than 10% missing
  mt_pre_norm_quot() %>% # Quotient Normalization
  mt_pre_trans_log() %>% # Log transform
  mt_pre_outlier_to_na() %>% # Detect outliers (any values that lie outside the 0.025 and 0.975 regions when assuming a normal distribution)  and set to NA
  mt_pre_impute_knn() %>% # Impute missing values with knn imputation
  mt_pre_batch_combat(batch_col = "batch_measured") # combat batch correction NEEDS TO BE DONE AFTER MISSING VALUES TAKEN CARE OF
```

Below we create a new column to specify the batches as apr21 vs apr25 and apr27 since this is how the data separates shown by the image above.

```{r}
# Make new dataframe
df_processed_not_standardized <- as.data.frame(t(assay(processed_SE_not_standardized)))
# Add columns from metadata
df_processed_not_standardized$Group <- processed_SE_not_standardized@colData$Group
df_processed_not_standardized$Organism <- processed_SE_not_standardized@colData$Organism
df_processed_not_standardized$date_measured <- processed_SE_not_standardized@colData$date_measured
df_processed_not_standardized$batch_measured <- processed_SE_not_standardized@colData$batch_measured
# Find the number of samples of each species that are remaining.
counts_organism_not_standardized <- table(df_processed_not_standardized$Organism)
# Display the result
print(counts_organism_not_standardized)
```
